Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics

### 2013-11-17 paper/paper2.tex
Many changes. The remaining thing is to write the prior work section

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\usepackage{soul}

@@

\section{Introduction}
Protein dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But atomic-resolution structural characterization of these dynamics is often a Herculean challenge, particularly in dynamical settings -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

Computer simulations can complement experiments by providing atomic-resolution insight into the structural dynamics. With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS}. In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field. The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}. For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Pande2010Everything, Beauchamp2012Simple, Prinz2011Markov, Bowman2013Quantitative}. Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics.

@@

Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics. To that end, we introduce a new procedure for scoring the likelihood of an MSM, which, together with cross validation and the Bayesian information criterion (BIC), enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states. This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables the treatment of model selection as an optimization problem amenable to automated methods.

@@

This emission distribution can be defined in a number of ways. The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics. When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume. This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}. For this reason, we can also choose other emission distributions that may be more tractable. In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings. The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering. If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.

@@

\begin{equation}
\label{eq:like_mvn}
\begin{split}
P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \sigma_{s(x_{i})}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \sigma_{s(x_{i})}^2}\right)
\end{split}
\end{equation} where $\mu_{s(x_i)}$ is the cluster center for the state that $x_i$ is assigned to. For simplicity, we follow \citet{Pelleg2000Xmeans} for estimating the intrastate variances, and simply use a maximum likelihood estimator so that the likelihood becomes:
\begin{equation}
\label{eq:like_mvn}
\begin{split}
P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
\end{split}
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
\end{equation} where $k$ is the number of states.

@@

Instead, we explore cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model. For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluate the likelihood on the left-out portion in an attempt to directly measure the generalization error. The BIC on the other hand involves augmenting the likelihood with a penalty on the number of parameters, and is
an asymptotic approximation Bayesian evidence.

@@

A further challenge is the procedure by which to model the volume of states which are at the ``edge'' of the MSM -- whose Voronoi cells extend to infinity in some direction. Is the volume of these states unbounded? It is appropriate to assert that the volume of these edge states is bounded in some way by the extent of our dataset. For example, the volume of a state might be defined as the volume of the intersection of its Voronoi cell and the convex hull of the whole dataset, which would encode the assumption that the likelihood of observing conformations outside the convex hull of the sampled data is vanishing.

We use a slightly modified version of this definition that adopts the same spirit. Instead of taking the outer bounding envelope to be the convex hull of the data, we take it to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$. For further efficiency, it is feasible to use only a random subsample of the dataset for this nearest neighbor computation.

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the first $10^6$ steps using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.

The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and furthermore, the likelihood remained almost exactly constant, which gives us some confidence that the volumes do not need to be perfect in order to have a good estimate of the likelihood.

For cross-validation, we used the final $10^6$ frames as the test data.

@@

As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate. The optimal models according to the three methods are between 200 and 1000 states for this system. This observation is consistent with the convergence of the timescales of the MSMs, but allows for a direct optimization of the state space, as opposed to a heuristic way of comparing models.

Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach. This could be due to the simplicity of this system. Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model. We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC. However, we note, that cross-validation requires estimating a model without some subset of the data, which is not feasible for large systems that are not over-sampled.

We also tested the MVN likelihood on MSMs built using k-means clustering on the M\"uller potential. The likelihood (\cref{fig:kmeans_mull}) peaks at a five-state model and actually decreases without the aid of the BIC or AIC. We can explain this maximum model by looking at where the cluster centers are located on the M\"uller potential. The five state model is the first model to have a cluster center located in the least stable well and as more states are added, the most stable well ends up being heavily split. We hypothesize that this over splitting in the most stable well gives rise to the decrease in likelihood.

@@

To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic}. In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements}, retaining between two and six uncorrelated linear combinations of residue-residue distances.

How many states are required for a Markov state model? Lane et. al., when analyzing this same dataset, proposed a 26,000 state Markov state model after clustering the data by RMSD, whereas Kellogg et. al. using an approach similar to that described herein, arrive at a \hl{1,000} state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}. Both likelihoods, however, tell us that while using tICA, we should build models that have 200-500 states (\cref{fig:ww}).

Since the MVN likelihood only requires the computation of distances, we can use it to analyze models built with in high dimensions as well as those built in spaces that are not vector spaces. We applied this methodology to models built using the minimum root mean square deviation in atom positions (RMSD) and k-centers clustering (\cref{fig:ww}). The MVN likelihood indicates that models built using RMSD should using on the order of 5,000 states, which is significantly higher than the tICA models. This is an illustration of how the tICA method can be used remarkably well to build Markov Models.

@@

According to this likelihood approach, the most likely models built on tICA projections are in the hundreds of states range.

For protein simulations, the slowest timescale (typically corresponding to folding) has been observed to be insensitive to differences in MSM construction methodology \cite{McGibbon2013Learning, others?}. We hypothesize that this behavior is due to the large structural changes associated with a folding event -- even a two state model, given the same dataset, could estimate the slowest timescale by simply counting the number of transitions. However, teasing out the slow timescales that are associated with subtler structural changes is a different challenge.

\hl{Kellogg stuff.}

@@

The likelihood described above allows for the comparison of MSMs with varying number of states, however it requires the definition of a vector space. As such, models built on different representations cannot be compared. This is because the likelihood requires the calculation of relative volumes, but the volumes will be different depending on what space you are working in. In addition, the lag time must be specified and cannot be selected using the likelihood. This is an improvement over the current state of constructing an MSM where the number of states is selected and then the lag time is tuned according to the eigenspectrum of the model. Selecting a lag time {\it a priori} is a much more intuitive parameter, that one typically knows based on the system.

We can also extend these results to models that do not have discrete states. Since all we require is an emission probability into a vector space, it is easy to use this likelihood for evaluating models that no longer have discrete states.

@@

Markov State Models are a promising analysis technique for automatically analyzing simulations generated from Markovian dynamics. There are a number of steps in the construction process, however, that require human intervention, which limits the use of MSMs to experts as well as introduces significant uncertainty into the model building process. These results take a large step toward fully automating the construction process by providing a quantitative way of selecting the number of states to use, as well as a way of selecting the clustering method. This will allow for non-experts to begin to use MSMs as well as allow adaptive sampling methods to be applied without human intervention.

>>>>

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external
                        % documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\usepackage{soul}

@@

\section{Introduction}
Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

Computer simulations can complement experiments by providing atomic insight into conformational dynamics. With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}. In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field. The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}. For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Pande2010Everything, Beauchamp2012Simple, Prinz2011Markov, Bowman2013Quantitative}. Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics. The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors to identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.

@@

Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics. To that end, we introduce a procedure for scoring the likelihood of an MSM, which, together with standard statistical model selection techniques, enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states. This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables the treatment of model selection as an optimization problem amenable to automated methods.

\section{Prior Work}
In the context of MSMs, model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.

\begin{itemize}
\item Swope-Pitera http://pubs.acs.org/doi/abs/10.1021/jp037421y
\item Noe http://www.pnas.org/content/106/45/19011
\item Park and Pande
\item Sergio
\item Kellogg
\end{itemize}

After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.

@@

This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model. We propose two possible distributions: first, one may model the conditional distribution $P(x_t | s(x_t))$ as a uniform distribution over the (hyper)volume of $s(x_t)$, as shown in eq. \ref{eq:like_vol}. This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation. A more tractable alternative is a Gaussian emission model where $P(x_t | s(x_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.  \ref{eq:like_vol} and used in the popular X-means algorithm.\cite{Pelleg2000Xmeans} The truncation error steps from the fact that the emission probability distributions from every state assign nonzero probability to each conformation in a dataset, while the parameterization procedure for the Markov transition probabilities assumes each conformation to be uniquely assigned to a single state. Relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.

%The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics. When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume. This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}. For this reason, we can also choose other emission distributions that may be more tractable. In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings. The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering. If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.

@@

\begin{equation}
\label{eq:like_mvn}
\begin{split}
&P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
    &\cdot \prod_{i=0}^T \left(2 \pi\right)^\frac{d}{2} |\Sigma_{s(x_i)}|^{-\frac{1}{2}}  \exp\left(-\frac{1}{2} (x_i - \mu_{s(x_i)})' \Sigma_{s(x_i)}^{-1} (x_i - \mu_{s(x_i)})\right)
\end{split}
\end{equation}

where $\mu_{s(x_i)}$ is the cluster center assigned to $x_i$ and $\Sigma_{s(x_i)}$ is the corresponding covariance matrix. For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which conditions eq. \ref{eq:like_mvn} reduces to
\begin{equation}
\label{eq:like_mvn2}
\begin{split}
P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
	&\cdot \prod_{i=0}^T \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
\end{split}
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
\end{equation} where $k$ is the number of states.

@@

Instead, we explore three alternative procedures: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model. In cross validation, we attempt to directly measure the Markov model's generalization error. We parameterize the model, building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, but evaluate the likelihood on the left-out portion. The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.

@@

A further challenge is the procedure by which to model the volume of states which are at the edge of the MSM -- whose Voronoi cells extend to infinity in some direction. In these cases the Voronoi cells are of unbounded volume. Because the convex hull of our simulation datasets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$.

For cross validation, we find it essential to use a dense prior during transition matrix estimation. Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes with maximum likelihood methods. For simplicity, we use the Jeffreys prior on each row of the transition matrix -- a Dirichlet prior with $\alpha=\frac{1}{2}$ -- which can be interpreted as a psuedocount of one half of an observed transition between each pair of states.

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{}. Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$. The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope. The second trajectory was used as a test set.

The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and the likelihood remained essentially constant, which indicates that the use of approximation volumes is sufficient to obtain robust estimates of the likelihood.

@@

As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate. The optimal models according to the three methods are between 200 and 1000 states for this system. This observation is consistent with the convergence of the relaxation timescales of the models.

Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach. This could be due to the simplicity of this system. Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model. We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC. However, we note, that cross-validation requires estimating a model without some subset of the data, which is not feasible for large systems in the data-sparse regime.

We also tested the Gaussian likelihood on MSMs built using k-means clustering on the M\"uller potential. The likelihood (\cref{fig:kmeans_mull}) peaks at a five-state model and decreases with the addition of further states. This maximum can be interpreted by inspecting the k-means cluster centers on the M\"uller potential. The five state model is the first model to have a cluster center located in the least stable well and as more states are added, the most stable well ends up being heavily split. We hypothesize that this over-splitting in the most stable minima gives rise to the decrease in MSM likelihood, and indicates a deficiency in clustering-based procedures which determine the state space without directly maximizing equations \ref{eq:like_vol} or \ref{eq:like_mvn}.

@@

To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic}. Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the the four slowest uncorrelated linear combinations of residue-residue distances. Using the Gaussian emission model, we compare both models built in the 3 dimensional tICA subspace and those built using k-centers clustering with the cartesian root-mean squared deviation (RMSD) over the Fip35 WW domain's 252 non-redundant heavy atoms, computed using the quaternion characteristic polynomial method\cite{Theobald2005Rapid}.

How many states are required in a Markov state model? The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by Shaw et. al.\cite{Shaw2010Atomic} have been used to parameterize multiple Markov state models. \citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas \citet{Kellogg2012Evaluation} clustering on contact maps, present a 1000 state model. We find the optimal number of states to be highly dependent on space in which the data is clustered. With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error. On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models: here we find high-quality models can be constructed using only on the order of 500 states. Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this dataset by \citet{McGibbon2013Learning}.

@@

@@

The likelihood functions described herein permits the comparison of Markov state models with varying number of states, however it requires the definition of a vector space over which the emission model is defined. As such, models built on different projections of an underlying space cannot be directly compared. Furthermore, this approach is not amenable to optimizing the Markov lag time, the discretization interval for the Markov jump process. Nonetheless, we anticipate that automatic selection of the number of states in a Markov state model will remove much of the guess work in MSM construction, as the selection of the lag time is more amenable to existing Chapman-Kolmogorov tests.

In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed.

@@

Markov State Models are powerful and popular tool for the analysis of molecular simulations, with a growing literature and two open source source software implementations\cite{Beauchamp2011Msmbuilder2, Senne2012EMMA}. There are, however a number of steps in the construction process that require human intervention, which limits the use of MSMs to experts and introduces significant biases into the model building process. Additionally, the ability to automatically construct Markov state models on the fly while simulations are in progress, an important point for so-called adaptive sampling procedures\cite{Bowman2010Enhanced}, is hampered when manual model selection is required. In the finite-data regime, additional complexity in the Markov model, in the form of increasing number of states, does come at a statistical cost.
These results take a step toward fully automating the MSM construction process and controlling this statistical error by providing a quantitative method for by which the most suitable discretization procedure can be selected.
